---
title: "Serbian Spa Waters Report"
author: "Leah Ye (300651931), Harry Philpott (300667756),


Claude Butler (300651409), Elizabeth Myers (300641471)"
output: 
  bookdown::pdf_document2:
    latex_engine: xelatex
bibliography: STAT394_references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE
)
set.seed(1234, kind="Mersenne-Twister")
library(dplyr)
library(knitr)
library(GGally)
library(ggplot2)
library(tidyr)
library(corrplot)
library(kableExtra)
library(plotly)
library(MASS)
```

# **Introduction**

This project is based on an article by Tanasković, Golobocanin, and Miljević [-@TANASKOVIC2012226], in which the authors analyzed hydrochemical data for mineral and thermal waters in Serbian territory. They used principal component analysis (PCA) and hierarchical cluster analysis (HCA) to characterize the chemical composition and natural radioactivity of these water samples, understand the relationships between the variables measured, and classify the waters with respect to their geotectonic units. PCA revealed four latent factors that accounted for 74.2% of the variability among the observations. Two of these four PCs are based on measures of mineralisation of the components of the host rock, and the other two are based on variables that are indicative of natural radioactivity. The results of PCA achieved a total correct classification of 83.3% for predefined geotectonic units, and the dendrogram of HCA classified the samples into four major groups and eleven subgroups.

# **Methodology**

The Serbian Spa Waters Dataset contains 30 observations (sites), with 15 columns. The data types and units are detailed in Table \@ref(tab:columnsTable).

```{r, echo = FALSE, results = 'asis'}
columns <- data.frame(
  Variable = c("ID", "Source", "geoStruct", "T", "pH", "EC", "TS",
               "Ca$^{2+}$", "Mg$^{2+}$", "Na$^{+}$", "K$^{+}$", "Cl$^{-}$", 
               "SO$^{2-}_4$", "HCO$^{-}_3$", "SiO$_2$"),
  Description = c("Index of each observation", "Name of the spa or spring",
                  "Geological structure", "Temperature", 
                  "pH level (Acidity/Alkalinity)", "Electrical conductivity",
                  "Total dissolved solids", "Calcium", "Magnesium",
                  "Sodium", "Potassium", "Chlorine", "Sulfate", "Bicarbonate",
                  "Silica, dissolved silicon dioxide"),
  Units = c("", "", "", "$^\\circ$C", "", "$\\mu$S/cm", "g/L", "mg/L", "mg/L",
            "mg/L", "mg/L", "mg/L", "mg/L", "mg/L", "mg/L"),
  DataType = c("Integer", "Character", "Integer", "Numerical", "Numerical", 
               "Integer", "Numerical", "Numerical", "Numerical", "Numerical",
               "Numerical", "Numerical", "Numerical", "Numerical", "Numerical"),
  stringsAsFactors = FALSE
)

kable(columns,
      caption = "Description of All Columns in the Dataset \\label{tab:columnsTable}",
      align = c("l", "l", "c", "l"),
      booktabs = TRUE,
      escape = FALSE) %>%  
  kable_styling(latex_options = c("HOLD_position"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE, extra_css = "font-family: serif;") %>%
  row_spec(1:nrow(columns), extra_css = "font-family: serif; line-height: 1.2;") %>%
  column_spec(1:ncol(columns), extra_css = "font-family: serif;")
```

Samples were taken from four different geological structures. The names of the structures and the integer values given in the dataset can be found in Table \@ref(tab:geoStructTable).

```{r, echo=FALSE, results='asis'}
basins <- data.frame(
  Integer = 1:4,
  Name = c("Hydrogeological basins", "Karstic terrains", 
           "Volcanogenic massifs", "Metamorphic regions"),
  stringsAsFactors = FALSE
)

kable(basins,
      caption = "Types of Geological Units \\label{tab:geoStructTable}",
      align = c("c", "l"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE, extra_css = "font-family: serif;") %>%
  row_spec(1:nrow(basins), extra_css = "font-family: serif; line-height: 1.2;") %>%
  column_spec(1:ncol(basins), extra_css = "font-family: serif;")
```


Ions $\mathrm{(HCO_3^-, Ca^{2+}, Mg^{2+}, Na^+, K^+, Cl^-, SO_4^{2-}, SiO_2)}$ 
were analysed in the laboratories of the Institute of Public Health of Serbia. 
Titration methods were applied to measure alkalinity of $Ca^{2+}$ and $Mg^{2+}$. 
Concentrations of $Na^+$ and $K^+$ were determined by atomic absorption 
spectrophotometry.

Six radiological variables from the original paper were not publicly available. This unfortunate omission does affect comparisons with those results, as we can not be sure what differences are due to our methodology and which are due to the reduced set of variables. The details of the missing variables are listed in Table \@ref(tab:isotopeTable).

```{r, echo=FALSE, results='asis'}
isotopes <- data.frame(
  Variable = c("GA", "GB", "$^{238}$U", "$^{228}$R", "$^{226}$R", "$^{40}$K"),
  Description = c("Gross alpha activities", "Gross beta activities", 
                  "Uranium isotope", "Radium isotope", "Radium isotope", "Potassium isotope"),
  Units = c("mBq/L", "", "", "", "", ""),
  stringsAsFactors = FALSE
)

kable(isotopes,
      caption = "Radioactive Isotopes Measured in Serbian Spa Waters \\label{tab:isotopeTable}",
      align = c("c", "l", "c"),
      booktabs = TRUE,
      escape = FALSE) %>%
  kable_styling(latex_options = c("HOLD_position"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE, extra_css = "font-family: serif;") %>%
  row_spec(1:nrow(isotopes), extra_css = "font-family: serif; line-height: 1.2;") %>%
  column_spec(1:ncol(isotopes), extra_css = "font-family: serif;")
```


Where Hyodrological Basins and Karstic terrains have 5 observations each, 
Volcanogen Massifs has 14 observations, and Metamorphic Regions has 6 
observations. This is an unbalanced dataset since there are varying observations 
per group.

There are no duplicates, missing data, zeros, negatives, NAs or NULL values in 
the dataset.

Exploratory data analysis will assess basic summary statistics of the data.
We will utilise graphical analysis of box plots and a QQ plot to 
check basic assumptions of  required for further analysis, such as both
univariate and multivariate normality. A correlation plot and Principal 
Component Analysis (PCA) will be performed to further investigate the 
relationships between variables. We will also identify outliers and 
surprising observations within the data, using standard deviation as well as
Mahalanobis distance, and in terms of the principal components.

Our initial exploratory data analysis revealed that the dataset was 
characterised by high variability and right-skewedness, leading us to perform
a log transformation on the data and use these transformed variables for our 
analysis.

We will assess the validity of the paper's geological structure groups based on the data, using the methods of MANOVA and Naive Bayes Classifier, as our EDA revealed that the data violates the assumptions for Linear Discriminant Analysis.

We will then use Factor Analysis to attempt to improve on the paper's model 
parsimony; factor analysis is better for model selection since it doesn't 
include noise like PCA, and maintains parsimony. 

# **Results and Discussion**

## EDA

```{r echo=FALSE}
data <- read.csv("serbianspawater.csv")
```

```{r echo=FALSE}
# Select numeric columns
numeric_cols <- data[sapply(data, is.numeric) & !(names(data) %in% c("idNum", "geoStruct"))]

# Create summary table
summary <- data.frame(
  Variable = names(numeric_cols),
  Mean     = round(sapply(numeric_cols, mean), 2),
  Median   = round(sapply(numeric_cols, median), 2),
  SD       = round(sapply(numeric_cols, sd), 2),
  Min      = round(sapply(numeric_cols, min), 2),
  Max      = round(sapply(numeric_cols, max), 2),
  row.names = NULL
)

# Format summary to show two decimal places
summary_formatted <- summary
summary_formatted[, 2:6] <- lapply(summary[, 2:6], function(x) sprintf("%.2f", x))

# Display table
kable(summary_formatted, 
      caption = "Summary Statistics for Numeric Columns", 
      align = c("l", "r", "r", "r", "r", "r"),
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("HOLD_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, extra_css = "font-family: serif;") %>%
  row_spec(1:nrow(summary_formatted), extra_css = "font-family: serif; line-height: 1.2;") %>%
  column_spec(1:ncol(summary_formatted), extra_css = "font-family: serif;")
```

This dataset is characterized by high variability. This is likely due to the 
different water sources. pH is more consistent, ranging from neutral (6.5) to 
slightly basic (9.0), which are fairly typical values for water in nature. The 
range of temperatures recorded in this dataset is from 14.40 to 95.00 degrees 
celcius. The large standard deviation confirms that there is a lot of 
variability in temperature. Some of these water sources are near boiling 
temperatures. Electrical conductivity also has a very large range, from 110.00 
to 5100.00 $\mu$S/cm. This indicates that some water samples have a lot of 
dissolved ions, leading to high conductivity. This measure is roughly 
proportional to the total dissolved solids, which also displays a large range, 
from 0.10 to 6.20. The maximum is more than three standard deviations above the 
mean, suggesting right-skew and surprising observations. There are eight major 
ions in this dataset: Calcium ($\mathrm{Ca^{2+}}$), 
Magnesium ($\mathrm{Mg^{2+}}$), Sodium ($\mathrm{Na^+}$), 
Potassium ($\mathrm{K^+}$), Chloride ($\mathrm{Cl^-}$), 
Sulfate ($\mathrm{SO_4^{2-}}$), Bicarbonate ($\mathrm{HCO_3^-}$), and 
Silica ($\mathrm{SiO_2}$). All of them are characterized by high variability, 
often with right-skew. Calcium has a mean of 61.02 and a range of 4.00 to 
178.00. The maximum is three standard deviations above the mean. Magnesium has 
an average of 44.46 and a range of 0.10 to 293.00, with a maximum that is four 
standard deviations above the mean. Sodium has a high average of 447.09, and a 
high maximum of 2280.00, which is again three standard deviations above the 
mean. Potassium is not as elevated with a mean of 16.58, but it is still 
somewhat variable. Chloride has a particularly extreme maximum of 1560.00, 
which is five standard deviations above its mean of 108.70. Sulfate has a mean 
of 57.53 and a maximum of 486.00, which is four standard deviations above the 
mean. Bicarbonate has a very high mean of 1297.91, and the maximum reaches a 
staggering 3538.00. The last major ion is Silica, which has a mean of 38.73, but 
a range from 2.00 to 120.00. To summarize, the whole dataset show high 
variability and pervasive right-skewness. The chemical composition of the water
samples is largely dominated by bicarbonate, with sodium and chloride also 
reaching very high values for some samples.

```{r boxplots, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.asp=0.8}
# Select numeric features for scaling 
X <- data[, 3:14]                    #all numerical catergories
X <- as.data.frame(scale(X))         # z-scores per variable
X$.row <- 1:nrow(X)

# reshaping is to long for ggplot so use tidyr library
box_long <- pivot_longer(           #tidyr
  X,
  cols = - .row,
  names_to = "variable",
  values_to = "value"
)

# boxplots on a common axis (z-scored)
ggplot(box_long, aes(x = variable, y = value)) +
  geom_boxplot(notch = TRUE, outlier.alpha = 0.6, na.rm = TRUE) +
  coord_flip() +
  labs(x = NULL, y = "Scaled value (z-score)",
       title = "Fig. 1: Boxplots of all variables (z-scored) for Serbian spa waters") +
  theme(panel.grid.minor = element_blank())
```
Variables were standardised to z-scores to allow direct comparison despite 
differing measurement units. The boxplots (Fig. 1) show that some features 
(e.g. $\mathrm{Na^+}$, $\mathrm{Cl^-}$, $\mathrm{HCO_3^-}$) have greater 
variability and distinct outliers, while others remain tightly clustered around 
the mean. This highlights which chemical measures are most heterogeneous across 
the spa waters. However, all variables in the boxplot have a non-symmetrical 
shape and contain outliers, indicating that a log transformation is necessary.

```{r multivariate normal, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.8}
# Fit MANOVA model using geoStruct as grouping
fit <- manova(as.matrix(X) ~ data$geoStruct)
# Extract residuals
res <- residuals(fit)
# Mahalanobis distances of residuals
mu_res <- colMeans(res)
S_res  <- cov(res)
d2_res <- mahalanobis(res, center = mu_res, cov = S_res)

# Degrees of freedom = number of variables
p <- ncol(res)

# QQ-plot
qqplot(qchisq(ppoints(nrow(res)), df = p), d2_res,
       main = "Fig. 2: QQ-plot of Serbian spa water MANOVA residuals",
       xlab = expression(paste(chi^2, " quantiles")),
       ylab = expression(paste("Ordered residual ", d[M]^2)))
abline(0,1,col="steelblue4")
```

The QQ-plot of MANOVA residual Mahalanobis distances against the theoretical 
$\chi^2$ distribution provides a diagnostic for assessing multivariate 
normality. If the residuals were perfectly normal, the points would lie close to 
the 45° line. In this case, the points deviate fairly evenly across the range of 
quantiles, rather than only at one end, indicating that the residuals are not an 
exact match to the multivariate Gaussian model. The upper tail in particular 
shows noticeable departures, with the largest residuals exceeding the expected 
$\chi^2$ quantiles, which indicates potential heavy-tailed behaviour or the 
presence of outliers. This means the assumption of multivariate normality is not 
fully satisfied.

```{r, echo=FALSE}
# Log transforming numeric variables and adding them into dataset as new columns
data$log.tempCels <- log(data$tempCels)
data$log.pH <- log(data$pH)
data$log.elec.Cond <- log(data$elecCond)
data$log.totSolid <- log(data$totSolid)
data$log.Ca2 <- log(data$Ca2)
data$log.Mg2 <- log(data$Mg2)
data$log.Na <- log(data$Na)
data$log.K <- log(data$K)
data$log.Cl <- log(data$Cl)
data$log.SO2 <- log(data$SO2)
data$log.HCO <- log(data$HCO)
data$log.SiO <- log(data$SiO)
```

After performing a log transformation of the variables, we recheck the 
univariate and multivariate normality of the data.

```{r echo=FALSE, warning=FALSE, message = FALSE}
X.log <- data[, 16:27]                    # all log transformed numerical variables
X.log <- as.data.frame(scale(X.log))         # z-scores per variable
X.log$.row <- 1:nrow(X.log)

box_long <- pivot_longer(           
  X.log,
  cols = - .row,
  names_to = "variable",
  values_to = "value"
)

# not z-scored
ggplot(box_long, aes(x = variable, y = value)) +
  geom_boxplot(notch = TRUE, outlier.alpha = 0.6, na.rm = TRUE) +
  coord_flip() +
  labs(x = NULL, y = "Scaled value (z-score)",
       title = "Fig.3: Boxplots of log transformed Serbian spa water variables (z-scored)") +
  theme(panel.grid.minor = element_blank())
```

The boxplots are far more symmetric than with the untransformed data, showing
more conformity with the normal distribution. However, we can clearly see the
presence of outliers, particularly with the variable $\mathrm{Mg^{2+}}$, with
four observations at least 2 standard deviations below the mean.

```{r echo=FALSE}
X_log_matrix <- as.matrix(X.log[,-which(names(X.log) == ".row")]) # Remove the .row column for analysis

# Fit MANOVA model
fit_log <- manova(X_log_matrix ~ data$geoStruct)
res_log <- residuals(fit_log)

# Reusing Prev code
# Mahalanobis distances of residuals
mu_res_log <- colMeans(res_log)
S_res_log  <- cov(res_log)
d2_res_log <- mahalanobis(res_log, center = mu_res_log, cov = S_res_log)

# Degrees of freedom = number of variables
p <- ncol(res_log)

# QQ-plot
qqplot(qchisq(ppoints(nrow(res_log)), df = p), d2_res_log,
       main = "Fig. 4: QQ-plot of MANOVA (log transformed) residuals",
       xlab = expression(paste(chi^2, " quantiles")),
       ylab = expression(paste("Ordered residual ", d[M]^2)))
abline(0,1,col="steelblue4")
```

The points follow the line much more closely, only slight deviations from the 
line at low and high values. The log transformed data much more closely 
follows a multivariate normal distribution.

```{r corrplot, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.8}
cor_mat <- cor(X, use = "pairwise.complete.obs", method = "pearson")

# Define layout: 2 rows, 1 column — small top row for title
layout(matrix(c(1, 2), nrow = 2), heights = c(0.1, 0.9))

# Top row: title only
par(mar = c(0, 0, 1, 0))  # No margins except top
plot.new()
title("Fig. 5: Correlation Matrix of Serbian Spa Water Variables", cex.main = 1.2)

corrplot(
  cor_mat,
  method = "color",        
  type = "upper",          
  order = "hclust",        
  addCoef.col = "black",   
  number.cex = 0.7,        
  tl.col = "black",        
  tl.srt = 45,             
  diag = FALSE,
  col = colorRampPalette(c("firebrick3", "white", "steelblue4"))(200)
)

``` 

Electrical conductivity, total solids, sodium, potassium, and chloride form a 
strongly correlated cluster, indicating they vary together across spa waters. 
In contrast, pH shows moderate negative correlations with several ions (e.g., 
C$\mathrm{Ca^{2+}}$, $\mathrm{Mg^{2+}}$, $\mathrm{HCO_3^-}$), suggesting 
different geochemical influences. Overall, the matrix highlights both tightly 
linked chemical groups and variables with weaker or opposing trends.


## PCA

```{r echo=FALSE, warning=FALSE, message=FALSE, PC-Elbow-Plot}

pca_fit <- prcomp(X_log_matrix, center=TRUE, scale. = TRUE) # apply pca on relevant explanatory numerical columns 

par(mar = c(5, 4, 4, 2) + 0.1)

# Plot PCA scree plot
plot(pca_fit, type = "l", main = "Fig. 7: PCA elbow plot for Serbian spa data")

```
```{r PCA-summary-table, echo=FALSE, message=FALSE, warning=FALSE}
# Perform PCA
pca_fit <- prcomp(X_log_matrix, center = TRUE, scale. = TRUE)

# Extract summary and convert to data frame
pca_summary <- summary(pca_fit)$importance
pca_df <- as.data.frame(round(pca_summary, 4))  # Round for readability
pca_df <- t(pca_df)  # Transpose so PCs are rows

# Create clean table
kable(pca_df, caption =  "Summary of Principal Components", align = "c", format = "latex",
      booktabs = TRUE,
      longtable = FALSE) %>%
  kable_styling(latex_options = c("H"), full_width = FALSE, position = "center") %>%
kable_styling(full_width = FALSE, position = "center") %>%
row_spec(0, extra_css = "font-family: serif;") %>%
row_spec(1:nrow(pca_df), extra_css = "font-family: serif;") %>%
column_spec(1:ncol(pca_df), extra_css = "font-family: serif;") 

```

Moving on to PCA, we can see (Table 2) that PC1 explains a lot of the variance 
(around half, 48.41%). PC2 explains about a further fifth (17.94%). PC3 and PC4
explain a further 10.83% and 7.79% respectively, and these 4 PCs cumulatively
account for 84.97% of the overall variance. From PC5 and onwards, each PC 
explains less than 7% of the extra variance. The elbow plot (Fig.6) displays this 
flattening as the effect of each PC becomes smaller and smaller. It appears 
that using 4 Principal Components would be sufficient in the model.

```{r echo=FALSE}

loadings <- pca_fit$rotation

loadings_df <- as.data.frame(loadings)
loadings_df$Variable <- rownames(loadings_df)

loadings_df <- loadings_df[, c("Variable", paste0("PC", 1:4))]

rownames(loadings_df) <- NULL

knitr::kable(loadings_df, digits = 3, caption = "Serbian Spa Variable Influence on Principal Components", align = "c")

```


In PC1, $HCO_3^-$ (0.388), $Na^+$ (0.345), elecCond (0.378), $K^+$ (0.376), 
and totSolid(0.367) have the highest influence. PC1 strongly correlates with 
variables relating to major ions and total dissolved content. PC1 measures the 
overall mineralisation and ionisation levels of the water. Samples with high PC1 
scores have high mineralisation. 

In PC2, $Mg^{2+}$ (0.438), $Ca^{2+}$ (0.55), tempCels(-0.329), $SiO_2$(-0.352),
and $Na^+$ (-0.301), have the strongest influence. PC2 separates water with high 
calcium and magnesium (hardness) from waters at higher temperatures. This could 
mean that there is a difference between hard, cold waters and warmer waters.

In PC3, $SO_2$ (0.717) and tempCels(0.521) have the strongest influence. Waters 
that have more sulfate are warmer.

In PC4, $SiO_2$ (-0.611), $Cl^-$(0.676) and tempCels(0.302) have the strongest 
influence. Waters with higher chlorine levels have less silicone and are warmer.

## Surprising observations

```{r echo=FALSE}
# Standardise numeric variables
X <- scale(data[, 3:14])

# Find any rows with at least one variable > 3 sd away
outliers <- which(apply(abs(X), 1, max) > 3)

# Show these "surprising" sites
surprising_sites <- data[outliers, c("waterSource", "tempCels", "elecCond", "totSolid",
                 "Na", "Cl", "HCO")]

knitr::kable(surprising_sites,
             format = "latex",
             caption = "Table: Water sources with at least one variable > 3 standard deviations from the mean",
             align = "c",
             digits = 2) %>%
  kable_styling(latex_options = "H", position = "center")
            
```

From the standardised $n=30$ observations, five springs are flagged as clear 
outliers: Junakovic, Selters, Vranjska, Tulaska, and Veluce. Each of these sites 
has at least one measurement more than three standard deviations from the mean. 
Junakovic is extreme in electrical conductivity ($5100~\mu$S/cm) and total 
dissolved solids ($6.2$ g/L). Selters is distinguished by exceptionally high 
concentrations of sodium ($2280$ mg/L), chloride ($1560$ mg/L), and bicarbonate 
($3360$ mg/L). Vranjska stands out for its near-boiling temperature 
($95^\circ$C), while Tulaska and Veluce show unusually high bicarbonate levels 
relative to the rest of the dataset. These five cases are the most “surprising” 
when the dataset is examined variable by variable.

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Multivariate check: Mahalanobis distances
mu <- colMeans(X)
S  <- cov(X)
d2 <- mahalanobis(X, center = mu, cov = S)

# Flag all sites with distance above the chi-square 99.9% cutoff
cut <- qchisq(0.999, df = ncol(X))
which(d2 > cut)
data[d2 > cut, c("waterSource", "tempCels", "elecCond", "totSolid",
                 "Na", "Cl", "HCO")]

# PCA on the log-transformed variables (already created earlier)
pca_result <- prcomp(X_log_matrix, scale. = TRUE)

# Get the scores for the first two PCs
scores <- as.data.frame(pca_result$x)
scores$waterSource <- data$waterSource
scores$geoStruct <- as.factor(data$geoStruct)

pca_plot_data <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  Geological_Group = as.factor(data$geoStruct),
  Water_Source = data$waterSource 
)

# PC2 vs PC1
ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Geological_Group)) +
  geom_point(size = 4, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  stat_ellipse(level = 0.95, alpha = 0.7) +  
  labs(
    title = "Fig. 8: PC2 vs PC1",
    x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "%)"),
    y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "%)"),
    color = "Geological\nStructure"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "right")
```
The PCA score plot gives a multivariate view, showing which sites are extreme 
in terms of their overall chemical profiles rather than a single ion. Along the 
first principal component (PC1, representing overall mineralisation), Ribarska 
and Prolom score strongly negative, indicating highly mineralised waters, while 
Kanjiža lies on the opposite end with a very dilute profile. Along the second 
principal component (PC2, contrasting hardness with temperature/pH), Vuča sits 
at the upper extreme with unusually high hardness, whereas Prolom and Ribarska 
again fall at the lower end, consistent with warmer, more alkaline waters. Thus, 
while the z-score analysis highlights individual extremes, the PCA identifies 
different sites as “surprising” because of their distinctive multivariate 
signatures.

## Factor Analysis

Factor analysis is a model estimation which deals with the assumptions of 
underlying causal structure, whereas PCA was a data transformation.


```{r echo=FALSE, message=FALSE}
library(psych)
X.log <- X.log[, !names(X.log) == ".row"] # remove '.row' column

pvalues <- rep(0, 5) 
for(f in 1:5) {
  res <- try(factanal(X.log, factors = f), silent = TRUE) # get the p-values for each factor
  if(!inherits(res, "try-error")) {
    pvalues[f] <- res$PVAL
  }
}

ggplot(data.frame(Factors = 1:5, pvalues = pvalues), 
       aes(x = Factors, y = pvalues)) +
  geom_point(size = 3) +
  geom_line() +
  xlab("Number of Factors") +
  ylab("p-value") +
  labs(title = "Fig. 9: Factor Adequacy Test")
```
The p-values in this table check the following:
$H_0:$ k factors are sufficient\
$H_1:$ More than k factors are needed

Therefore, 5 factors is sufficient as p-value is \> 0.05. 4 or less factors is 
not sufficient (p-values \< 0.03).

```{r warning=FALSE, message=FALSE, echo=FALSE}
# Run factor analysis
faresult <- factanal(X.log, factors = 5)

# Extract SS loadings (sum of squared loadings per factor)
ss_loadings <- colSums(as.matrix(faresult$loadings)^2)

# Calculate proportions and cumulative proportions
proportion_var <- ss_loadings / ncol(X.log)
cumulative_var <- cumsum(proportion_var)

# Combine into a matrix for display
stats_matrix <- rbind(
  "SS loadings"    = ss_loadings,
  "Proportion Var" = proportion_var,
  "Cumulative Var" = cumulative_var
)

# Round for readability
stats_matrix <- round(stats_matrix, 3)

# Assign column names like Factor1, Factor2, etc.
colnames(stats_matrix) <- paste0("Factor", seq_len(ncol(stats_matrix)))

knitr::kable(stats_matrix, "latex", booktabs = TRUE, caption = "Summary Statistics from Factor Analysis (5 Factors)") %>%
  kable_styling(latex_options = "hold_position")

```

Factor 5's loading is small (0.379), and only contributes 3.1% of the overall
variance. So we will go back to four factors, and visualize the result.

```{r echo=FALSE}
fa.diagram(faresult$loading[,1:4], digits=4, main = "")
title(main = "Fig. 10: Factor analysis" )
```

The red arrows indicate negative variables.

-   F1: Total Solid (+), Electrical Conductivity (+), Na (+), HCO (+), K (+), 
SiO (+).\
-   F2: Mg2 (+), Ca2 (+), pH (-), tempCels (-).\
-   F3: Cl (+).\
-   F4: SO2 (+).

Factor analysis selected 4 factors with 79% variance explained, whereas PCA 
preferred 4 PCs with 85% variance explained. The factors show that there are 4 
water types:

-   F1: Highly mineralised waters (high soliditity and electrical conductivity)\
-   F2: Magenesium, Calcium (hardness ions), acidic and cool waters\
-   F3: Chloric waters.\
-   F4: Sulfate waters.

Factor analysis is better for model selection since it doesn't include noise 
like PCA (that's why PCA variance for 4 PCs was greater than FA 4 factors 
cumulative variance), and maintains parsimony. PCA had to consider all 12 
Principal Components, whereas FA only had to look at 5 or less factors for 
determining the underlying structure of spa water composition.

## geoStruct Groups

```{r echo=FALSE}
# Check p value
summary(fit_log, test = "Hotelling-Lawley")
```
To assess whether the predefined geostructural groups are verified by the water 
chemistry data, we began with a formal test of multivariate mean differences. A 
MANOVA on the log-transformed variables produced a p-value of 0.057, which 
provides only weak evidence that the group mean vectors differ. This suggests 
that while some differences exist, they are not pronounced enough to be 
considered statistically significant at conventional levels.

```{r  echo=FALSE}
lda_data <- data.frame(geoStruct = data$geoStruct, X_log_matrix)
lda_fit <- lda(geoStruct ~ ., data = lda_data)
lda_scores <- predict(lda_fit)$x

plot_data <- data.frame(lda_scores, Group = as.factor(lda_data$geoStruct))

ggplot(plot_data, aes(x = LD1, y = LD2, color = Group)) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.95) +
  labs(title = "Fig. 6: Visual Check of Group Covariance",
       x = "Linear Discriminant 1",
       y = "Linear Discriminant 2") +
  theme_minimal()
```

We can see from the LDA plot distinct shapes and orientations in ellipses 
representing the covariance of the groups. Given the clear differences between
these groups, the covariance matrices of the groups cannot be equal and LDA is 
not appropriate for this data. The assumption of equal covariance is clearly 
failed when looking at the distinction in the shape and orientation between each 
group.

As an alternative, we adopted the Naive Bayes classifier. Unlike LDA, Naive Bayes does not require estimation of full covariance matrices and instead assumes conditional independence of the variables within each class. This makes it more appropriate for datasets with small and unbalanced groups such as ours, where some geostructural categories contained as few as five observations. We fitted the Naive Bayes model to the log-transformed variables and evaluated performance using leave-one-out cross-validation to provide a fair estimate of classification accuracy.

Although we did not have external prior probabilities for the geostructural categories, Naive Bayes estimates them from the observed class frequencies in the sample. This corresponds to assuming that the sample is representative of the population proportions. Alternatively, one could impose equal priors to reflect an uninformative prior belief. In either case, the classification rule remains valid, and our results reflect how well the chemical profiles align with the predefined groups under these assumptions.

```{r echo=FALSE}
library(e1071)
library(caret)
y  <- factor(data$geoStruct)
Xl <- scale(as.data.frame(X_log_matrix))
# Leave-One-Out Cross-Validation
n <- nrow(Xl); pred <- factor(rep(NA,n), levels=levels(y))
for(i in 1:n){
  fit_i <- naiveBayes(x=Xl[-i,], y=y[-i])
  pred[i] <- predict(fit_i, newdata=Xl[i,,drop=FALSE])
}
confusionMatrix(table(True=y, Pred=pred))

```

The model achieved an overall accuracy of 50 percent, which is higher than the no-information rate of 40 percent but still modest. The Kappa statistic of 0.29 suggests only fair agreement between the predicted and true classes. The classification performance varied across groups. Karstic terrains (Class 2) and Metamorphic regions (Class 4) were the best recognised categories, with balanced accuracies of 0.81 and 0.74 respectively, indicating that their chemical profiles are relatively distinctive. Hydrological Basins (Class 1), on the other hand, were poorly distinguished and often misclassified as Volcanogen Massifs (Class 3). Volcanogen Massifs themselves, despite being the largest group, were only moderately well classified, with a balanced accuracy of 0.53.

These results suggest that the groups, the geostructural categories, are only partially verified by the data. Some groups do exhibit clear multivariate chemical signatures, allowing them to be separated reasonably well, while others show substantial overlap that prevents reliable classification. In particular, the difficulty in separating Hydrological Basins from Volcanogen Massifs implies that the water chemistry of these groups is not sufficiently distinct. Overall, the analysis provides evidence that geological structure does influence chemical composition, but the effect is not strong enough across all groups for the classification to be considered fully verified. The geostructural groupings are therefore only partially confirmed by the data.




