---
title: "Serbian Spa Waters Report"
author: "[GUYS ADD YOUR NAMES AND STUDENT ID HERE], Leah Ye 300651931"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# **Introduction & Background** - Liz

Hydrochemical data for the mineral and thermal waters studied on Serbian territory.

# **Descriptive analysis of the data** - Leah

The Serbian Spa Waters Dataset contains 30 observations(sites), with 15 columns:  

* idNum: Integer - Index of each observation;  
* waterSource: Character - Spa/Spring the observation belongs to;     
* tempCels: Numerical - Outflowing temperature (°C);
* pH: Numerical - pH level(Acidity/Alkalinity);  
* elecCond: Integer - Electrical Conductivity of water measured immediately after collection on the spot, total ion concentration (μS/cm);   
* totSolid: Numerical - Total dissolved solids (gL)
* Ca2: Numerical -  Calcium (mg/L);  
* Mg2: Numerical - Magnesium (mg/L);    
* Na: Numerical - Sodium (mg/L);   
* K: Numerical - Potassium (mg/L);   
* Cl: Numerical - Chlorine (mg/L);  
* S02: Numerical - Sulfate (mg/L);   
* HCO: Numerical - Bicarbonate (mg/L);    
* Si0: Numerical - Silica, dissolved silicon dioxide (mg/L);    
* geoStruct: Integer - The geological structure the water was collected from (1 = Hydrological Basin, 2 = Karstic terrains, 3 = Volcanogen Massifs, 4 = Metamorphic Regions)

Where Ions$(HCO_3^−, Ca^{2+}, Mg^2+, Na^+, K^+, Cl^−, SO^{2−}_4, SiO_2)$ were analysed in the laboratories of Instutute of Public Health of Serbia. Titration methods were applied to measure alkalinaty of $Ca^{2+}$ and $Mg^{2+}$. Concentrations of $Na^+$ and $K^+$ were determined by atomic absorption spectrophotometry.  

Where Hyodrological Basins and Karstic terrains have 5 observations each, Volcanogen Massifs has 14 observations, and Metamorphic Regions has 6 observations. 

## **Summary Statistics** - Liz

```{r echo=FALSE}
data <- read.csv("serbianspawater.csv")
```



# **Graphical analysis** 

## Pairs Plot - Liz

## Box Plot - Harry

## Correlation matrix - Harry

## Eigenvalues - Claire

## PCA - Leah

# Multivariate Normal - Harry

# Surprising Observations - Claire




# Factor Analysis

# geoStruct groups


