---
title: "Serbian Spa Waters Report"
author: "[GUYS ADD YOUR NAMES AND STUDENT ID HERE], Leah Ye 300651931, Harry Philpott 300667756"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(knitr)
library(GGally)
library(ggplot2)
library(tidyr)
library(corrplot)
```


# **Introduction & Background** - Liz

Hydrochemical data for the mineral and thermal waters studied on Serbian territory.

# **Descriptive analysis of the data** - Leah

The Serbian Spa Waters Dataset contains 30 observations(sites), with 15 columns:  

* idNum: Integer - Index of each observation;  
* waterSource: Character - Spa/Spring the observation belongs to;     
* tempCels: Numerical - Outflowing temperature (°C);
* pH: Numerical - pH level(Acidity/Alkalinity);  
* elecCond: Integer - Electrical Conductivity of water measured immediately after collection on the spot, total ion concentration (μS/cm);   
* totSolid: Numerical - Total dissolved solids (gL)
* Ca2: Numerical -  Calcium (mg/L);  
* Mg2: Numerical - Magnesium (mg/L);    
* Na: Numerical - Sodium (mg/L);   
* K: Numerical - Potassium (mg/L);   
* Cl: Numerical - Chlorine (mg/L);  
* S02: Numerical - Sulfate (mg/L);   
* HCO: Numerical - Bicarbonate (mg/L);    
* Si0: Numerical - Silica, dissolved silicon dioxide (mg/L);    
* geoStruct: Integer - The geological structure the water was collected from (1 = Hydrological Basin, 2 = Karstic terrains, 3 = Volcanogen Massifs, 4 = Metamorphic Regions)

Where Ions$(HCO_3^−, Ca^{2+}, Mg^2+, Na^+, K^+, Cl^−, SO^{2−}_4, SiO_2)$ were analysed in the laboratories of Institute of Public Health of Serbia. Titration methods were applied to measure alkalinity of $Ca^{2+}$ and $Mg^{2+}$. Concentrations of $Na^+$ and $K^+$ were determined by atomic absorption spectrophotometry.  

Where Hyodrological Basins and Karstic terrains have 5 observations each, Volcanogen Massifs has 14 observations, and Metamorphic Regions has 6 observations.  

There are no duplicates, missing data, zeros, negatives, NAs or NULL values in the dataset. 

## **Summary Statistics** - Liz

```{r echo=FALSE}
data <- read.csv("serbianspawater.csv")
```

```{r}
# Select numeric columns
numeric_cols <- data[sapply(data, is.numeric)]

# Create summary table
summary <- data.frame(
  Variable = names(numeric_cols),
  Mean     = round(sapply(numeric_cols, mean), 2),
  Median   = round(sapply(numeric_cols, median), 2),
  SD       = round(sapply(numeric_cols, sd), 2),
  Min      = round(sapply(numeric_cols, min), 2),
  Max      = round(sapply(numeric_cols, max), 2),
  row.names = NULL
)

# Display table
kable(summary, caption = "Summary Statistics for Numeric Columns")
```

# **Graphical analysis** 

## Pairs Plot - Liz

```{r echo=FALSE}
# Create pairs plot 
pairs_plot <- ggpairs(
  numeric_cols,
  upper = list(continuous = wrap("cor", size = 6, color = "black")), 
  lower = list(continuous = wrap("points", alpha = 0.6, size = 2)), 
  diag  = list(continuous = wrap("densityDiag", alpha = 0.5))        
) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12, color = "black"),
    axis.text.y = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 14, color = "black"),
    strip.text = element_text(size = 14, color = "black")
  )

# Save as PNG
ggsave("full_pairs_plot_points_bigger.png", plot = pairs_plot, width = 25, height = 25, units = "in", dpi = 300)
```


## Box Plot - Harry
```{r boxplots, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.asp=0.8}


# Select numeric features for scaling 
X <- data[, 3:14]                    #all numerical catergories
X <- as.data.frame(scale(X))         # z-scores per variable
X$.row <- 1:nrow(X)

# reshaping is to long for ggplot so use tidyr library
box_long <- pivot_longer(           #tidyr
  X,
  cols = - .row,
  names_to = "variable",
  values_to = "value"
)

# boxplots on a common axis (z-scored)
ggplot(box_long, aes(x = variable, y = value)) +
  geom_boxplot(notch = TRUE, outlier.alpha = 0.6, na.rm = TRUE) +
  coord_flip() +
  labs(x = NULL, y = "Scaled value (z-score)",
       title = "Serbian spa water: Boxplots of all variables (z-scored)") +
  theme(panel.grid.minor = element_blank())
```
Variables were standardised to z-scores to allow direct comparison despite differing measurement units. The boxplots show that some features (e.g. Na⁺, Cl⁻, HCO₃⁻) have greater variability and distinct outliers, while others remain tightly clustered around the mean. This highlights which chemical measures are most heterogeneous across the spa waters.



## Correlation matrix - Harry

```{r corrplot, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.8}
cor_mat <- cor(X, use = "pairwise.complete.obs", method = "pearson")
corrplot(cor_mat,
         method = "color",        
         type = "upper",          
         order = "hclust",        
         addCoef.col = "black",   
         number.cex = 0.7,        
         tl.col = "black",        
         tl.srt = 45,             
         diag = FALSE, 
         col = colorRampPalette(c("firebrick3","white","steelblue4"))(200)
        
)
```
Electrical conductivity, total solids, sodium, potassium, and chloride form a strongly correlated cluster, indicating they vary together across spa waters. In contrast, pH shows moderate negative correlations with several ions (e.g., Ca²⁺, Mg²⁺, HCO₃⁻), suggesting different geochemical influences. Overall, the matrix highlights both tightly linked chemical groups and variables with weaker or opposing trends.


## Eigenvalues - Claire

## PCA - Leah
```{r echo=FALSE, warning=FALSE, message=FALSE, PC-Elbow-Plot, fig.cap="Figure 4: PCA plot", fig.asp=0.8}
library(plotly)
pca_fit <- prcomp(data[3:14], center=TRUE, scale. = TRUE) # apply pca on relevant explanatory numerical columns 
plot(pca_fit, type="l")
```

* PC1 explains most of the variance (around 4)  
* PC2 explains a fair part of the variance
* At PC3 and 4, the extra variance explained is a bit less  
* From PC5, the line starts to flatten  
* Each additional component after PC5 only adds a tiny bit of variance. 

```{r echo=FALSE, PC-Summary-tab}
summary(pca_fit)
```

* PC1 has a standard deviation of 2.05, and explains 35.1% of the variance
* PC2 adds 17.6%, and explains 52.6% of the variance
* PC3 adds 12.4%, and explains 65.0% of the variance  
* PC4 adds 9.7%, and explains 74.7% of the variance
* After PC4 onwards, each PC explains less than 9% of the extra variance
* PC5 explains 83.2% of the variance
* PC6 explains 90.5% of the variance
* PC9 explains the most variance (99.1%)

```{r}
pca_fit
```

* 

# Multivariate Normal - Harry
```{r multivariate normal, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.asp=0.8}
# Fit MANOVA model using geoStruct as grouping
fit <- manova(as.matrix(X) ~ data$geoStruct)
# Extract residuals
res <- residuals(fit)
# Mahalanobis distances of residuals
mu_res <- colMeans(res)
S_res  <- cov(res)
d2_res <- mahalanobis(res, center = mu_res, cov = S_res)

# Degrees of freedom = number of variables
p <- ncol(res)

# QQ-plot
qqplot(qchisq(ppoints(nrow(res)), df = p), d2_res,
       main = "QQ-plot of MANOVA residuals",
       xlab = expression(paste(chi^2, " quantiles")),
       ylab = expression(paste("Ordered residual ", d[M]^2)))
abline(0,1,col="steelblue4")
```
The QQ-plot of MANOVA residual Mahalanobis distances against the theoretical $\chi^2$ distribution provides a diagnostic for assessing multivariate normality. If the residuals were perfectly normal, the points would lie close to the 45° line. In this case, the points deviate fairly evenly across the range of quantiles, rather than only at one end, indicating that the residuals are not an exact match to the multivariate Gaussian model. The upper tail in particular shows noticeable departures, with the largest residuals exceeding the expected $\chi^2$ quantiles, which indicates potential heavy-tailed behaviour or the presence of outliers. Strictly speaking, this means the assumption of multivariate normality is not fully satisfied. Nevertheless, since most observations adhere reasonably closely to the expected distribution, it is common in practice to proceed under the approximate normality assumption while acknowledging these limitations and interpreting MANOVA results with some caution.


# Surprising Observations - Claire




# Factor Analysis

# geoStruct groups


